Life in the canopy: community trait assessments reveal substantial functional diversity among fern epiphytes

### 2019-10-27 ms/manuscript.Rmd
Split intro into one sentence per line
Update MS
Revise intro

The appearance of angiosperm-dominated forests in the late Cretaceous and early Cenozoic had an immense effect on terrestrial ecosystems [@Niklas1983; @Kenrick1997a; @Lloyd2008]. The heterogeneous niche space of the angiosperm canopy comprising stratified light and humidity levels and a range of substrates for growth became rapidly filled by a variety of epiphytic plants, including mosses [@Fiz-Palacios2011], leafy liverworts [@Feldberg2014], orchids [@Givnish2015b], bromeliads [@Givnish2014a], and ferns [@Schneider2004d]. Diversification rates in some of these groups may have further increased during the Paleocene / Eocene thermal maximum, a period of warmer and wetter climates that may have enabled movement into the canopy [@Schuettpelz2009; @Testo2016a]. The transition to epiphytic growth has been particularly important for fern diversity, with ca. 29% of all species exhibiting this growth form relative to ca. 10% across all vascular plants [@Kress1986]. Major transitions to epiphytism occurred at least five times during the evolution of ferns, each followed by diversification during the late Cretaceous or early Cenozoic [@Schuettpelz2009; @WatkinsJr.2012]. Recent studies [@Sundue2015; @Testo2016a; @Lehtonen2017] have failed to find evidence of rapid diversification characteristic of adaptive radiations associated with epiphytic growth in ferns, but have confirmed gradually increasing rates of diversification in Polypodiaceae [@Sundue2015] and generally support the significance of epiphytic growth as a key innovation for extant fern diversity (@Testo2016a; @WatkinsJr.2012; but see @Lehtonen2017).

Despite the ecological opportunities afforded by life in the canopy, including escape from disturbance [@Page2002] and abundant light for photosynthesis [@Benzing2000], epiphytes must contend with greater stress in the form of increased evaporation and extreme oscillations in temperature relative to terrestrial plants [@Zotz2001; @Lowman2012]. Associated with this change, epiphytic ferns have evolved various morphological adaptations that minimize water loss from their leaves, including shorter stipes [@WatkinsJr.2010], thicker leaves [@Kluge2007; @WatkinsJr.2007b], and protection with scales or hairs [@WatkinsJr.2006b]. Some lineages have specialized adaptations to epiphytic growth, such as humus-collection (e.g., *Asplenium*; @Benzing1990, *Drynaria*; @Janssen2005), vegetative desiccation tolerance (DT) (e.g., Hymenophyllaceae; @Shreve1911, *Pleopeltis*; @Pessin1925), CAM photosynthesis (e.g., *Pyrrosia*; @Wong1976), and symbiotic relationships with ants (e.g., *Antrophyum*; @WatkinsJr.2008, *Lecanopteris*; @Gay1993a). Many of these same adaptations are found in epiphytic seed plants [@Benzing1990]. However, the evolution of epiphytism in ferns was more complex given that ferns rely on a free-living gametophyte [@Tsusumi2013]. The fern gametophyte is anatomically simple, and differs from the sporophyte in anatomy, morphology, and physiology. Such differences create a system of 'dual ecology' where selective pressures acting on one stage are different than those acting on another. For example, DT is prevalent in gametophytes of multiple lineages of epiphytic ferns, but is rare in sporophytes [@WatkinsJr.2007c]. Despite evidence for adaptations to epiphytism in multiple lineages in both sporophytes and gametophytes, no studies to our knowledge have analyzed the evolution of such traits across the fern phylogeny including both life stages and representatives of all five Cretaceous epiphytic radiations *sensu* [@Schuettpelz2009].

Most studies investigating adaptations to epiphytism have utilized a univariate, cross-species approach, analyzing traits individually between terrestrial and epiphytic species [e.g., @WatkinsJr.2007b; @WatkinsJr.2010]. However, given multiple radiations into the epiphytic niche [@Schuettpelz2009], it is likely that traits related to epiphytism carry at least some degree of phylogenetic signal (i.e., the tendency for closely related species to resemble each other). Therefore, a comparative approach incorporating phylogenetic relationships (and if possible, branch lengths) is appropriate [@Felsenstein1985]. However, such approaches have only recently been applied to studies of epiphytism in ferns, and at relatively limited phylogenetic scales [e.g., @Zhang2014b; @Sundue2015].

It should be possible to gain greater insight into the selective strength exerted by the epiphytic habitat by utilizing a community-based approach. If the harsh abiotic conditions of the canopy act as a filter, we expect to find lower functional richness at a community level, which should result in decreased phylogenetic diversity under a scenario of phylogenetic trait conservatism [@Webb2002]. However, the relative strength of this environmental filter may depend on overall abiotic conditions; for example, relative humidity (RH) and temperature also vary with elevation [@Cardelus2006]. Elevation is a strong predictor of fern species richness, with maximal richness occurring at mid-elevations, especially in the tropics [@Kessler2011a]. Functional diversity is also correlated with species richness (few species tend to have exactly the same functional traits; @Schleuter2010). Thus, environmental variation along elevational gradients is expected to impact the functional, phylogenetic, and species diversity of epiphytic and terrestrial fern communities.

Here, we investigate the evolution and role in community assembly of functional traits related to epiphytic growth in fern gametophytes and sporophytes on Moorea, French Polynesia. Moorea is an isolated tropical oceanic island with a high phylogenetic diversity of ferns distributed over a steep elevational gradient, making it ideal for a community phylogenetic analysis of functional traits. Specifically, we test the following hypotheses related to epiphytic growth in ferns:

1. The epiphytic environment experiences greater extremes in temperature and humidity than the terrestrial environment, thus acting as a filter on the traits of plants that grow there.

2. Epiphytic ferns have converged upon similar morphologies multiple times in both sporophyte and gametophyte phases.

3. Epiphytic fern communities have lower functional and phylogenetic diversity relative to terrestrial communities due to environmental filtering, and the strength of this filtering varies with changes in environment.

>>>>

The appearance of angiosperm-dominated forests in the late Cretaceous and early Cenozoic fundamentally changed climates and had an immense effect on terrestrial ecosystems, including increased global precipitation and the creation of novel niches [@Niklas1983; @Kenrick1997a; @Lloyd2008; @Boyce2010]. 
The heterogeneous niche space of the angiosperm canopy comprising stratified light and humidity levels and a range of substrates for growth became rapidly filled by a variety of epiphytic plants, including mosses [@Fiz-Palacios2011], leafy liverworts [@Feldberg2014], orchids [@Givnish2015b], bromeliads [@Givnish2014a], and ferns [@Schneider2004d]. 
The transition to epiphytic growth was particularly important for modern fern diversification: ca. 29% of all fern species are epiphytes vs. ca. 10% across all vascular plants [@Kress1986].
Moreover, major transitions to epiphytism occurred at least five times within ferns, each followed by diversification during the late Cretaceous or early Cenozoic [@Schuettpelz2009; @WatkinsJr.2012].
This has led to the hypothesis that epiphytism was a key factor contributing to extant fern diversification [@Schneider2004d; @Schuettpelz2009; @WatkinsJr.2012].
Recent studies exploring this idea [@Sundue2015; @Testo2016a; @Lehtonen2017], however, have largely failed to identify evidence of rapid diversification associated with epiphytism in ferns. 
This raises the possibility that modern fern diversification was not shaped by the evolution of epiphytism *per se*, but rather by the exploitation of various niches within the canopy and on the forest floor.

To better understand the role of the canopy in fern diversification, one obvious direction is to explore putatively adaptive fern epiphyte traits in a phylogenetic context. 
In contrast to most terrestrial plants, epiphytes must contend with greater stress in the form of nutrient limitation, increased evaporative demand, drought, and extreme oscillations in temperature  [@Zotz2001; @WatkinsJr.2009; @Lowman2012]. 
Previous studies have identified multiple traits as likely adaptations to these conditions, albeit many without consideration of phylogeny. 
Epiphytic ferns tend to have morphological features that minimize water loss from their leaves, including shorter stipes [@WatkinsJr.2010], thicker leaves [@Kluge2007; @WatkinsJr.2007b], and protection with scales or hairs [@WatkinsJr.2006b]. 
Additional specialized adaptations to epiphytic growth include humus collection (e.g., *Asplenium*; @Benzing1990, *Drynaria*; @Janssen2005), vegetative desiccation tolerance (DT) (e.g., Hymenophyllaceae; @Shreve1911, *Pleopeltis*; @Pessin1925), CAM photosynthesis (e.g., *Pyrrosia*; @Wong1976), and symbiotic relationships with ants (e.g., *Antrophyum*; @WatkinsJr.2008, *Lecanopteris*; @Gay1993a).

While many similar adaptations are found in epiphytic seed plants and serve as key reference points  [@Benzing1990], the evolution of epiphytism in ferns is more complex owing to their unique life-cycle. 
Ferns produce a free-living gametophyte that is typically smaller than the sporophyte and relies on aquatic motile sperm.
Unlike the sporophyte, the fern gametophyte lacks vascular tissue, stomata, or cuticle, placing it in constant equilibrium with its external environment. 
In parallel with some of the adaptations in the sporophytes, the gametophytes of some epiphytic ferns also exhibit morphological and physiological adaptations to minimize water loss and tolerate water scarcity. 
For example, extreme desiccation tolerance is prevalent in gametophytes of multiple lineages of epiphytic ferns, but is rare in sporophytes  [@WatkinsJr.2007c; Pittermann2013].
In addition, epiphytic fern gametophytes tend to exhibit more complex three-dimensional morphologies that may function to retain water and slow its loss [WatkinsJr.2007c].
Despite evidence for adaptations to epiphytism in multiple lineages in both sporophytes and gametophytes, no previous studies to our knowledge have analyzed the functional ecology of fern epiphytes in a phylogenetic context including both life stages and representatives of all five Cretaceous epiphytic radiations *sensu* [@Schuettpelz2009].

To gain insight into the role of the canopy in promoting or restricting diversity, a functional and community-based approach can be applied to study this question. 
If the canopy is uniformly harsher than the terrestrial environment, thus acting as a "filter", we would expect functional convergence of epiphytes at the community level, which in turn may result in decreased phylogenetic diversity of epiphytic communities if traits are phylogenetically conserved [@Webb2002; @Cavender-Bares2009]. 
However, the relative strength of this environmental filter may depend on overall abiotic conditions; for example, relative humidity (RH) and temperature also vary with elevation [@Cardelus2006]. 
Conversely, the stratification of the canopy may stimulate functional diversification by niche partitioning, leading to increased functional diversity of epiphytic communities and reduced phylogenetic signal of relevant traits. 
Furthermore, the role of the canopy as a promoter or restrictor of diversity may be modulated by overall abiotic conditions, including most notably relative humidity (RH) and temperature. 
For instance, elevation is a strong predictor of fern species richness, with maximal richness occurring at mid-elevations, especially in the tropics [@Kessler2011a]. 
Functional diversity is also correlated with species richness [@Schleuter2010]. 
Thus, environmental variation along elevational gradients is expected to impact the functional, phylogenetic, and species diversity of epiphytic and terrestrial fern communities.

Here, we seek to understand the role of the canopy in fern diversification by investigating traits related to epiphytic growth in fern gametophytes and sporophytes on Moorea, French Polynesia.
Moorea is an isolated tropical oceanic island with high phylogenetic diversity of ferns distributed across a steep elevational gradient, making it ideal for a community phylogenetic analysis of functional traits. 
Specifically, we test the following hypotheses related to epiphytic growth in ferns:

1. Epiphytic ferns have converged upon similar morphologies in both sporophyte and gametophyte phases.

2. Epiphytic fern communities have lower functional and phylogenetic diversity relative to terrestrial communities.
